\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-09 paper/ms.tex
Some minor edits.

To start, we define a rank-$R$ semiseparable matrix as a matrix $K$ where the
elements can be written as
\begin{eqnarray}
K_{n,m} = \left\{\begin{array}{ll}
    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
    A_{n,n} & \mathrm{if}\,m=n \\
    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
N$-dimensional diagonal matrix $A$.
This definition can be equivalently written as
\begin{eqnarray}\eqlabel{semi-sep}
K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
\end{eqnarray}
where the $\mathrm{tril}$ function extracts the strict lower triangular
component of its argument and $\mathrm{triu}$ is the equivalent strict upper
triangular operation.
In this representation, rather than storing the full $N \times N$ matrix $K$,
it is sufficient to only store the $N \times R$ matrices $U$ and $V$, and the
$N$ diagonal entries of $A$, for a total storage volume of $(2\,R+1)\,N$
floating point numbers.
Given a semiseparable matrix of this form, it is known that matrix-vector
products and matrix-matrix products can be evaluated in $\mathcal{O}(N)$
operations and the interested reader is directed to \citet{Vandebril:2007} for
more details about semiseparable matrices in general.

@@

\response{We select about 180~days of contiguous observations of
KIC~1430163 from \kepler.
This dataset has \exampleiiindata\ measurements and, using a tuned linear
algebra
implementation\footnote{We use the Intel Math Kernel Library
\url{https://software.intel.com/en-us/intel-mkl}}, a
single evaluation of the likelihood requires over 8~seconds on a modern Intel
CPU.
This calculation, using \celerite\ with the model in \eq{rot-kernel}, only
takes $\sim 1.5\,\mathrm{ms}$~--~a speed-up of more than three orders of
magnitude per model evaluation.

@@

We run \exampleiiinburn~steps of burn-in, followed by \exampleiiinsteps~steps
of MCMC using \emcee.
We estimate the integrated autocorrelation time of the chain for $\ln
P_\mathrm{rot}$ and estimate that we have \exampleiiineff\ independent
samples.
These samples give a posterior constraint on the period of $P_\mathrm{rot} =
\rotationperiod$ and the marginalized posterior distribution for $P$ is shown
in the right panel of \Figure{rotation}.
This result is in good agreement with the literature value with smaller
uncertainties.
A detailed comparison of GP rotation period measurements and the traditional
methods is beyond the scope of this paper, but forthcoming work (R.~Angus,
\etal, submitted) demonstrates that GP inferences are, at a population level,
more reliable than other methods.
}

@@

\response{
For this demonstration, we randomly select a month-long segment of PDC
\citep{Stumpe:2012, Smith:2012} \kepler\ data.
Unlike the previous examples, the posterior distribution is sharply
multimodal and na\"ively maximizing the posterior using \project{L-BFGS-B} is
not practical.
Instead, we start by estimating the initial values for $\nu_\mathrm{max}$ and
$\Delta\nu$ using only the month-long subset of data and following the
standard procedure described by \citet{Huber:2009}.
We then run set of \project{L-BFGS-B} optimizations with values of
$\nu_\mathrm{max}$ selected in logarithmic grid centered on our initial
estimate of $\nu_\mathrm{max}$ and initial values of $\Delta\nu$ computed
using the empirical relationship between these two quantities
\citep{Stello:2009}.
This initialization procedure is sufficient for this example, but the general
application of this method will require a more sophisticated prescription.

@@

\response{
The continuous autoregressive moving average (CARMA) models introduced into
the astrophysics literature by \citet{Kelly:2014} share many features with
\celerite.
CARMA models are derived by solving a stochastic differential equation and,
like \celerite, the likelihood function for a CARMA model conditioned on a
one-dimensional dataset can be solved in $\mathcal{O}(N)$.
The kernel function for a CARMA$(J,\,K)$ model is \citep{Kelly:2014}
\begin{eqnarray}\eqlabel{carma-kernel}
k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{carma-coeff}
A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
\end{eqnarray}
and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
the model.
Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
every CARMA model corresponds to an equivalent \celeriteterm\ model and the
parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
The inverse statement is not as simple.
In practice, this means that \celerite\ could be trivially used to compute any
CARMA model.
Using the CARMA solver to compute a \celeriteterm\ model, however, requires
solving \eq{carma-coeff} numerically for a given set of $\{A_j\}_{j=1}^J$.
While the computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$,
the Kalman filtering based method for solving CARMA models \citep{Kelly:2014}
is, in practice, somewhat faster and the memory requirements are smaller.
One benefit of the \celeriteterm\ model is that it is parameterized in terms
of the amplitudes directly.
This means that the physical interpretation of these parameters is, in some
cases, simpler.
On the other hand, CARMA models are parameterized by modeling the power
spectrum of the stochastic driving force and this might be the quantity of
interest in other situations.
Another benefit of the \celeriteterm\ framework is that, unlike
filtering-based CARMA solvers, it directly solves matrix equations of the form
${K_\bvec{\alpha}}^{-1}\,b$ in $\mathcal{O}(N)$.
This is a crucial feature when GP modeling is combined with linear models
\citep[for example][]{Luger:2017}.
}

@@

Many other approximate methods for scaling GP inference exist \citep[see, for
example,][and references therein]{Wilson:2015a} and we make no attempt to make
our discussion exhaustive.
The key takeaway here is that \celerite\ provides an \emph{exact} method for
GP inference for a specific, but flexible, class of one-dimensional kernel
functions.
Furthermore, since \celeriteterm\ models can be interpreted as a mixture of
stochastically-driven, damped simple harmonic oscillators, they are a
physically motivated choice of covariance function in many applications.

@@

We evaluate the coefficients of each of the polynomial in the series by
evaluating $f_0 = \{Q_0(0),\,\ldots,\,Q_{2\,(J-1)+1}(0)\}$ to give us the signs
of these polynomials evaluated at $z=0$.
Likewise, we evaluate the coefficients of the largest order term in each
polynomial that gives the sign of the polynomial as $z \rightarrow \infty$,
$f_\infty$.
With the sequence of coefficients $f_0$ and $f_\infty$, we then determine how
many times the sign changes in each of these, where $\sigma(0)$ is the number
of sign changes at $z=0$, and $\sigma(\infty)$ is the number of sign changes
at $z \rightarrow \infty$.
The total number of real roots in the range
$(0,\,\infty]$ is given by $N_{+}=\sigma(0)-\sigma(\infty)$.

@@

\section{Semiseparable operations}

To help clarify these equations, we write out here an example of a semiseparable matrix with a
single \celeriteterm\ term with $c=d=1$: %and $N=3$
\begin{eqnarray}\eqlabel{semi-separable-structure}
K&=&%
\begin{pmatrix}
%    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-c\,\tau_{2,1}}\left(a\,\cos{(d \tau_{2,1})}+b\,\sin{(d \tau_{2,1})}\right)} & {\scriptscriptstyle e^{-c\,\tau_{3,1}}\left(a\,\cos{(d \tau_{3,1})}+b\,\sin{(d \tau_{3,1})}\right)}\\
%    {\scriptscriptstyle e^{-c\,\tau_{2,1}}\left(a\,\cos{(d \tau_{2,1})}+b\,\sin{(d \tau_{2,1})}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)}\\
%    {\scriptscriptstyle e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)}&{\scriptscriptstyle  e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)} & \scriptscriptstyle{a + {\sigma_3}^2}
%    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)} & {\scriptscriptstyle e^{-\tau_{3,1}}\left(a\,\cos{\tau_{3,1}}+b\,\sin{\tau_{3,1}}\right)}\\
%    {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}\\
%    {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}&{\scriptscriptstyle  e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2}
    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)} & {\scriptscriptstyle e^{-\tau_{31}}\left(a\,\cos{\tau_{31}}+b\,\sin{\tau_{31}}\right)} & \ldots\\
    {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \ldots\\
    {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)}&{\scriptscriptstyle  e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2} & \ldots \\
    \ldots & \ldots & \ldots & \ldots
\end{pmatrix}\\
&=&
\begin{pmatrix}
%    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} \\
%    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} \\
%    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2}
    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & \ldots\\
    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} & \ldots\\
    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2} & \ldots\\
%    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & \ldots \\
%    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& \ldots \\
    \ldots & \ldots & \ldots & \ldots
\end{pmatrix}\\
&+&\mathrm{tril}\left[\begin{pmatrix}
%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\cos{(d t_1)}+b\,\sin{(d t_1)}\right)}&
%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\sin{(d t_1)}-b\,\cos{(d t_1)}\right)}\\
%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\cos{(d t_2)}+b\,\sin{(d t_2)}\right)}&
%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\sin{(d t_2)}-b\,\cos{(d t_2)}\right)}\\
%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\cos{(d t_3)}+b\,\sin{(d t_3)}\right)}&
%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\sin{(d t_3)}-b\,\cos{(d t_3)}\right)}
    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}\\
   \ldots & \ldots
\end{pmatrix} %\right.\\
%&\times &
\times
%\left.
\begin{pmatrix}
% {\scriptscriptstyle e^{c\,t_1} \cos{(d t_1)}} &{\scriptscriptstyle e^{c\,t_2} \cos{(d t_2)}} & {\scriptscriptstyle e^{c\,t_3} \cos{(d t_3)}}\\
% {\scriptscriptstyle e^{c\,t_1} \sin{(d t_1)}} & {\scriptscriptstyle e^{c\,t_2} \sin{(d t_2)}} & {\scriptscriptstyle e^{c\,t_3} \sin{(d t_3)}}
% {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_3} \cos{t_3}}\\
% {\scriptscriptstyle e^{t_1} \sin{t_1}} & {\scriptscriptstyle e^{t_2} \sin{t_2}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}
 {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
 {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
 {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}\\
 \ldots & \ldots
\end{pmatrix}^T\right]\\
&+&
\mathrm{triu}\left[\begin{pmatrix}
% {\scriptscriptstyle e^{c\,t_1} \cos{(d t_1)}} & {\scriptscriptstyle e^{c\,t_1} \sin{(d t_1)}} \\
% {\scriptscriptstyle e^{c\,t_2} \cos{(d t_2)}} & {\scriptscriptstyle e^{c\,t_2} \sin{(d t_2)}}\\
% {\scriptscriptstyle e^{c\,t_3} \cos{(d t_3)}} & {\scriptscriptstyle e^{c\,t_3} \sin{(d t_3)}}
 {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
 {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
 {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}\\
 \ldots & \ldots
\end{pmatrix} % \right.\\
\times
%\left.
\begin{pmatrix}
%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\cos{(d t_1)}+b\,\sin{(d t_1)}\right)}&
%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\cos{(d t_2)}+b\,\sin{(d t_2)}\right)}&
%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\cos{(d t_3)}+b\,\sin{(d t_3)}\right)}\\
%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\sin{(d t_1)}-b\,\cos{(d t_1)}\right)}&
%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\sin{(d t_2)}-b\,\cos{(d t_2)}\right)}&
%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\sin{(d t_3)}-b\,\cos{(d t_3)}\right)}
%    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
%    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
%    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}\\
%    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}&
%    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}&
%    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}
    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}\\
   \ldots & \ldots
\end{pmatrix}^T\right]\\
&=& A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T).
\end{eqnarray}

@@

Simulation of a Gaussian process for large $N$ requires multiplying an
$N$-dimensional vector of Gaussian normal deviates, $\bvec{q}$, by the square root
of the covariance matrix:
\begin{equation}
\bvec{y} = K_\bvec{\alpha}^{1/2} \bvec{q},
\end{equation}
where $q_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that $\mathcal{Q} = -2 \mathcal{L}$
with $\mu_\bvec{\theta} =0$ is given by
\begin{eqnarray}
\mathcal{Q} &=&\bvec{y}^T K_\bvec{\alpha}^{-1} \bvec{y}\\
&=& \bvec{y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} \bvec{y}\\
&=& \bvec{q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} \bvec{q}\\
&=& \bvec{q}^T \bvec{q}.
\end{eqnarray}
The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $\mathcal{Q} \sim \chi_N^2$.

>>>>

To start, we define a rank-$R$ semiseparable matrix as a matrix $K$ where the
elements can be written as
\begin{eqnarray}
K_{n,m} = \left\{\begin{array}{ll}
    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
    A_{n,n} & \mathrm{if}\,m=n \\
    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
N$-dimensional diagonal matrix $A$.
This definition can be equivalently written as
\begin{eqnarray}\eqlabel{semi-sep}
K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
\end{eqnarray}
where the $\mathrm{tril}$ function extracts the strict lower triangular
component of its argument and $\mathrm{triu}$ is the equivalent strict upper
triangular operation.  An example of this type of matrix is given in appendix \ref{operations}.
In this representation, rather than storing the full $N \times N$ matrix $K$,
it is sufficient to only store the $N \times R$ matrices $U$ and $V$, and the
$N$ diagonal entries of $A$, for a total storage volume of $(2\,R+1)\,N$
floating point numbers.
Given a semiseparable matrix of this form, it is known that matrix-vector
products and matrix-matrix products can be evaluated in $\mathcal{O}(N)$
operations and the interested reader is directed to \citet{Vandebril:2007} for
more details about semiseparable matrices in general.

@@

\response{We select about 180~days of contiguous observations of
KIC~1430163 from \kepler.
This dataset has \exampleiiindata\ measurements and, using a tuned linear
algebra
implementation\footnote{We use the Intel Math Kernel Library
\url{https://software.intel.com/en-us/intel-mkl}}, a
single evaluation of the likelihood requires over 8~seconds on a modern Intel
CPU.
This calculation, using \celerite\ with the model in \eq{rot-kernel}, only
takes $\sim 1.5\,\mathrm{ms}$~--~a speed-up of more than three orders of
magnitude per model evaluation. % :) closer to four orders of magnitude!

@@

We run \exampleiiinburn~steps of burn-in, followed by \exampleiiinsteps~steps
of MCMC using \emcee.
We estimate the integrated autocorrelation time of the chain for $\ln
P_\mathrm{rot}$ and estimate that we have \exampleiiineff\ independent
samples.
These samples give a posterior constraint on the period of $P_\mathrm{rot} =
\rotationperiod$ days and the marginalized posterior distribution for $P$ is shown
in the right panel of \Figure{rotation}.
This result is in good agreement with the literature value with smaller
uncertainties.
A detailed comparison of GP rotation period measurements and the traditional
methods is beyond the scope of this paper, but forthcoming work (R.~Angus,
\etal, submitted) demonstrates that GP inferences are, at a population level,
more reliable than other methods.
}

@@

\response{
For this demonstration, we randomly select a month-long segment of PDC data
\citep{Stumpe:2012, Smith:2012} \kepler\ data.
Unlike the previous examples, the posterior distribution is sharply
multimodal and na\"ively maximizing the posterior using \project{L-BFGS-B} is
not practical.
Instead, we start by estimating the initial values for $\nu_\mathrm{max}$ and
$\Delta\nu$ using only the month-long subset of data and following the
standard procedure described by \citet{Huber:2009}.
We then run set of \project{L-BFGS-B} optimizations with values of
$\nu_\mathrm{max}$ selected in logarithmic grid centered on our initial
estimate of $\nu_\mathrm{max}$ and initial values of $\Delta\nu$ computed
using the empirical relationship between these two quantities
\citep{Stello:2009}.
This initialization procedure is sufficient for this example, but the general
application of this method will require a more sophisticated prescription.

@@

\response{
The continuous autoregressive moving average (CARMA) models introduced into
the astrophysics literature by \citet{Kelly:2014} share many features with
\celerite.
CARMA models are derived by solving a stochastic differential equation and,
like \celerite, the likelihood function for a CARMA model conditioned on a
one-dimensional dataset can be solved in $\mathcal{O}(N)$.
The kernel function for a CARMA$(J,\,K)$ model is \citep{Kelly:2014}
\begin{eqnarray}\eqlabel{carma-kernel}
k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{carma-coeff}
A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
\end{eqnarray}
and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
the model; $\sigma$ is real, while the others are complex.
Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
every CARMA model corresponds to an equivalent \celeriteterm\ model and the
parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
The inverse statement is not as simple.
In practice, this means that \celerite\ could be trivially used to compute any
CARMA model.
Using the CARMA solver to compute a \celeriteterm\ model, however, requires
solving \eq{carma-coeff} numerically for a given set of $\{A_j\}_{j=1}^J$.
The computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$ using
the Kalman filtering based method for solving CARMA models \citep{Kelly:2014};
however, in practice this method is somewhat slower.
An additional benefit of the \celeriteterm\ model is that it is parameterized in terms
of the amplitudes directly.
This means that the physical interpretation of these parameters is, in some
cases, simpler.
On the other hand, CARMA models are parameterized by modeling the power
spectrum of the stochastic driving force and this might be the quantity of
interest in other situations.
Another benefit of the \celeriteterm\ framework is that, unlike
filtering-based CARMA solvers, it directly solves matrix equations of the form
${K_\bvec{\alpha}}^{-1}\,b$ in $\mathcal{O}(N)$.
This is a crucial feature when GP modeling is combined with linear models
\citep[for example][]{Luger:2017}.
}

@@

Many other approximate methods for scaling GP inference exist \citep[see, for
example,][and references therein]{Wilson:2015a} and we make no attempt to make
our discussion exhaustive.
The key takeaway here is that \celerite\ provides an \emph{exact} method for
GP inference for a specific, but flexible, class of one-dimensional kernel
functions.
Furthermore, since \celeriteterm\ models can be interpreted as a mixture of
stochastically-driven, damped simple harmonic oscillators, they are a
physically motivated choice of covariance function in many astronomical applications.

@@

We evaluate the coefficients of each of the polynomials in the series by
evaluating $f_0 = \{Q_0(0),\,\ldots,\,Q_{2\,(J-1)+1}(0)\}$ to give us the signs
of these polynomials evaluated at $z=0$.
Likewise, we evaluate the coefficients of the largest order term in each
polynomial that gives the sign of the polynomial as $z \rightarrow \infty$,
$f_\infty$.
With the sequence of coefficients $f_0$ and $f_\infty$, we then determine how
many times the sign changes in each of these, where $\sigma(0)$ is the number
of sign changes at $z=0$, and $\sigma(\infty)$ is the number of sign changes
at $z \rightarrow \infty$.
The total number of real roots in the range
$(0,\,\infty]$ is given by $N_{+}=\sigma(0)-\sigma(\infty)$.

@@

\section{Semiseparable operations} \label{operations}

To help clarify equations \ref{system}, we write out here an example of a semiseparable matrix with a
single \celeriteterm\ term with $c=d=1$:
\begin{eqnarray}\eqlabel{semi-separable-structure}
K&=&%
\begin{pmatrix}
    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)} & {\scriptscriptstyle e^{-\tau_{31}}\left(a\,\cos{\tau_{31}}+b\,\sin{\tau_{31}}\right)} & \ldots\\
    {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \ldots\\
    {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)}&{\scriptscriptstyle  e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2} & \ldots \\
    \ldots & \ldots & \ldots & \ldots
\end{pmatrix}\\
&=&
\begin{pmatrix}
    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & \ldots\\
    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} & \ldots\\
    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2} & \ldots\\
    \ldots & \ldots & \ldots & \ldots
\end{pmatrix}\\
&+&\mathrm{tril}\left[\begin{pmatrix}
    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}\\
   \ldots & \ldots
\end{pmatrix}
\times
\begin{pmatrix}
 {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
 {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
 {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}\\
 \ldots & \ldots
\end{pmatrix}^T\right]\\
&+&
\mathrm{triu}\left[\begin{pmatrix}
 {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
 {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
 {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}\\
 \ldots & \ldots
\end{pmatrix} 
\times
\begin{pmatrix}
    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}\\
   \ldots & \ldots
\end{pmatrix}^T\right]\\
&=& A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T).
\end{eqnarray}

In addition to solving $K^{-1} {\bf y}$, several other operations can be
carried out recursively in $\mathcal{O}(N)$, simulation, multiplication, 
and prediction, which we describe in the following sections.

@@

Simulation of a Gaussian process for large $N$ requires multiplying an
$N$-dimensional vector of Gaussian normal deviates, $\bvec{q}$, by the square root
of the covariance matrix:
\begin{equation}
\bvec{y} = K_\bvec{\alpha}^{1/2} \bvec{q},
\end{equation}
where $q_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the first term of $\mathcal{Q} = -2 \mathcal{L}$
with $\mu_\bvec{\theta} =0$ is given by
\begin{eqnarray}
\mathcal{Q} &=&\bvec{y}^T K_\bvec{\alpha}^{-1} \bvec{y}\\
&=& \bvec{y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} \bvec{y}\\
&=& \bvec{q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} \bvec{q}\\
&=& \bvec{q}^T \bvec{q}.
\end{eqnarray}
The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $\mathcal{Q} \sim \chi_N^2$; for
diagonal $K_\bvec{\alpha}$, $\mathcal{Q}$ is the $\chi^2$ statistic.
